Towards automated provenance collection for experimental runs of agent-based models

### 2023-07-20 paper/paper.tex
Pre initial submission version
Updated paper for ESSA2023 reviewer comments

%\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
% \author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
% Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }

%%% GP: to discuss -- which other authors should we include here?
%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
%
%\authorrunning{D. Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
% \institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
% \email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
%\url{https://www.hutton.ac.uk/}
%}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

Replication of social simulation results has been highlighted as a significant
issue for the agent-based modelling (ABM) community for a number of years (e.g. \cite{edmonds2003replication}).
The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
the analysis of the outputs of the model can potentially be just as complex a
process as developing and running the model itself. Analysis of outputs is no less difficult
to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also 
many scientific workflow tools, such as Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
these are workflow tools, the focus is on automation and
repeatability rather than provenance, which, if it is included, is as an afterthought.

@@

Each of the Bash commands listed in table \ref{tab:commands} composes over % What does 'composes over' mean?
the above Python commands to populate the \texttt{SSREPI} database in a consistent and
logical manner. Broadly speaking, {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
\ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
\ttvar{SSREPI_input}}, {\color{blue} \ttvar{SSREPI_output}} and {\color{blue}
\ttvar{SSREPI_argument}} are responsible for recording coarse grain provenance.
{\color{blue} \ttvar{SSREPI_value}}, {\color{blue}
\ttvar{SSREPI_visualisation_variable_value}}, {\color{blue}
\ttvar{SSREPI_statistical_variable_value}}, {\color{blue} \ttvar{SSREPI_run}}
and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
remaining primitives are largely about recording other metadata. Further information may be found
in the tool's public repository \footnote{\url{https://github.com/DougSalt/ABM-metadata}}.

@@

Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in the
paper \cite{polhill2017lessons}, showing some of the workflow and provenance graphs, and then using fictitious examples to apply the workflow and provenance metadata to addressing issues in large-scale experiments. From a workflow perspective, the main result is that the
principal diagrams in that paper were successfully recreated, with the same, albeit
not identical, results. %right? % Plus, I've deleted all the whining :-)

@@

        \item \textbf{Finegrain} - a provenance diagram down to the level of
        variables.  \item \textbf{Folksonomomy} - a diagram showing annotations
            against the database, produced and categorised at the discretion of
            the user doing the annotation \item \textbf{Project} - management
            metadata. Largest granularity of metadata supported \item
            \textbf{Provenance} - provenance diagram at the level of file and
        parameter \item \textbf{Services} - service provided and requirement
            description \item \textbf{Workflow} - the actual workflow
\end{itemize}

The supplementary programs use the Dot language for input to Graphviz\cite{ellson2002graphviz}, producing the diagrams such as those in figures
\ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
\ref{fig:high-lit-provenance-trace}. The workflow graph is in
fig. \ref{fig:workflow}. The resultant provenance graph is massive, since there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
\ref{fig:sub-provenance}.

@@

So what use is this provenance metadata? Imagine for the purposes of illustration that
we have found a bug in a script. The entity {\color{red}\ttvar{Applications.application_3831436655}} is an experiment setup script \ttvar{SSS-StopC2-Cluster-create.pl}. We can use the workflow graph to see what else in the workflow might be affected. As might be expected for an experiment setup script, there are serious cascading consequences, which we can visualise in Figure \ref{fig:broken-application}. The Dot language used by Graphviz constitutes a primitive graph
database. Indeed there are programs that can transform Dot files into
TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. To generate Figure \ref{fig:broken-application}, we used the workflow visualisation Dot file rather than the relational database.

@@

>>>>

%\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
%
\titlerunning{An automated provenance collection framework}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Doug Salt\inst{1} \orcidID{0000-0001-5186-9388} \and
 Gary Polhill\inst{1} \orcidID{0000-0002-8596-0590} \and
 Corran Musk\inst{1} \and
 Lorenzo Milazzo\inst{2} \and
 Dawn Parker\inst{3} \orcidID{0000-0001-8988-193X}  }

%%% GP: to discuss -- which other authors should we include here?
%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
%
\authorrunning{D. Salt et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
\url{https://www.hutton.ac.uk/}
\and Satellite and Surface Assimilation (SSA) Group
Met Office, FitzRoy Road, Exeter, Devon, EX1 3PB, UK
\email{lorenzo.milazzo@physics.org}
\and University of Waterloo, 200 University Avenue West
Waterloo, ON, Canada  N2L 3G1
\url{https://uwaterloo.ca/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.

@@

Replication of social simulation results has been highlighted as a significant
issue for the agent-based modelling (ABM) community for a number of years (e.g. \cite{edmonds2003replication}).
The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
the analysis of the outputs of the model can potentially be just as complex a
process as developing and running the model itself. Analysis of outputs is no less difficult
to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also 
many scientific workflow tools, such as Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
these are workflow tools, the focus seems to be  on automation and
repeatability rather than provenance, which, if it is included, seems not to be the primary purpose of the described tools. There are exceptions to this, such as the open provenance framework \cite{moreau2008open}, implemented in \cite{gadelha2011provenance} with yet another workflow control langauge \cite{zhao2007swift}, but such tools do not seem to have flourished and seemingly for agent-based models are non existent.

@@

Each of the Bash commands listed in table \ref{tab:commands} composes over % What does 'composes over' mean?
the above Python commands to populate the \texttt{SSREPI} database in a consistent and
logical manner. Broadly speaking, {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
\ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
\ttvar{SSREPI_input}}, {\color{blue} \ttvar{SSREPI_output}} and {\color{blue}
\ttvar{SSREPI_argument}} are responsible for recording coarse grain provenance.
{\color{blue} \ttvar{SSREPI_value}}, {\color{blue}
\ttvar{SSREPI_visualisation_variable_value}}, {\color{blue}
\ttvar{SSREPI_statistical_variable_value}}, {\color{blue} \ttvar{SSREPI_run}}
and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
remaining primitives are largely about recording other metadata. Further information may be found
in the tool's public repository. %\footnote{\url{https://github.com/DougSalt/ABM-metadata}}.

@@

Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in
the paper \cite{polhill2017lessons}, showing some of the workflow and
provenance graphs, and then using fictitious examples to apply the workflow and
provenance metadata to addressing issues in large-scale experiments. From a
workflow perspective, the main result is that the principal diagrams in that
paper were successfully recreated, with the same, albeit not identical,
results. %right? % Plus, I've deleted all the whining :-)
% whoops the reviewer complained about there being nothing here, so I am going
% to have to reintrouduce some of the whining. :)
By this we mean that the diagrams produced for the original paper
\cite{polhill2017lessons} were "roughly the same". They were not identical, but
contained similar features close enough to illustrate the features that were originally reported on in the original research from
\cite{polhill2017lessons}.

@@

        \item \textbf{Finegrain} - a provenance diagram down to the level of
        variables.  \item \textbf{Folksonomy} - a diagram showing annotations
            against the database, produced and categorised at the discretion of
            the user doing the annotation \item \textbf{Project} - management
            metadata. Largest granularity of metadata supported \item
            \textbf{Provenance} - provenance diagram at the level of file and
        parameter \item \textbf{Services} - service provided and requirement
            description \item \textbf{Workflow} - the actual workflow
\end{itemize}

The supplementary programs use the Dot language for input to Graphviz\cite{ellson2002graphviz}, producing the diagrams such as those in figures
\ref{fig:workflow}, \ref{fig:sub-provenance}, \ref{fig:broken-application} and
\ref{fig:high-lit-provenance-trace}. The workflow graph is in
fig. \ref{fig:workflow}. The resultant provenance graph is massive, since there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
\ref{fig:sub-provenance}.

@@

So what use is this provenance metadata? Imagine for the purposes of illustration that
we have found a bug in a script. The entity {\color{red}\ttvar{Applications.application_3831436655}} is an experiment setup script \ttvar{SSS-StopC2-Cluster-create.pl}. We can use the workflow graph to see what else in the workflow might be affected. As might be expected for an experiment setup script, there are serious cascading consequences, which we can visualise in Figure \ref{fig:broken-application}. The Dot language used by Graphviz constitutes a primitive graph
database. Indeed there are programs that can transform Dot files into TinkerPop
GraphSON format \cite{tinkerpop2015graphson,gv2graphson}, two reasonably well known and utilised graphing database formats. To generate Figure
\ref{fig:broken-application}, we used the workflow visualisation Dot file
rather than the relational database.

@@

\section{Discussion}

It might seem that we are confusing random stream control and reproducibility
of individual simulation runs, i.e., given a set of inputs and random stream
that the simulation would always produce the same results, versus the
reproducibility of an overall analysis that comes out of a calibration or other
simulation workflow. However, the latter issues cannot even begin to be
addressed until a reproducible and reliable workflow is implmented for agent based models -
that is a directed acyclic graph from inputs to published outputs/visualisation
and/or results. This may seem like a contradiction to our earlier claim that
this is not a workflow tool, but workflow in agent-based models is complex and
the link from data/model to published material for agent-based models is
currently missing. Agent based modelling does not have any specialist needs in
terms of work control other than magnitude on the number of inputs, number of
runs and number of resuts. However step ups in magnitude are not always trivial
as the step up petaflop to exaflop is demonstrates
\cite{north2008agent}.
